Reviving stochasticity: eccentricity convergence depends on black hole binary orbit

### 2023-05-09 papers/paper-1/diagrams/hm_coresersic.tex

    % define nodes
    % define layer 1
    \node[obs] (Shat) {$\hat{\Sigma}$};
    \factor[below=of Shat] {S_f} {right:Lognorm} {} {};
    \node[det, below=of S_f] (S) {$\Sigma$};
    \node[obs, below=of S] (R) {$\hat{R}$};
    \node[obs, right=of S] (sigma) {$\hat{\sigma_\Sigma}^2$};

    % define layer 2
    \node[latent, left=of S] (Sb) {$\tilde{\Sigma}_\mathrm{b}$};
    \node[latent, below=of Sb] (Re) {$R_\mathrm{e}$};
    \node[latent, below=of Re] (rb) {$r_\mathrm{b}$};
    \node[latent, above=of Sb] (g) {$\gamma$};
    \node[latent, above=of g] (n) {$n$};
    \node[latent, above=of n] (a) {$a$};

    % define layer 3
    \factor[left=of rb] {rb_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of rb_f, yshift=-1.2em] (rbs) {$r_{\mathrm{b},\sigma}$};
    \node[hyper, left=of rb_f, yshift=1.2em] (rbm) {$r_{\mathrm{b},\mu}$};

    \factor[left=of Re] {Re_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of Re_f, yshift=-1.2em] (Res) {$R_{\mathrm{e},\sigma}$};
    \node[hyper, left=of Re_f, yshift=1.2em] (Rem) {$R_{\mathrm{e},\mu}$};

    \factor[left=of Sb] {Sb_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of Sb_f, yshift=-1.2em] (Sbs) {$\tilde{\Sigma}_{\mathrm{b},\sigma}$};
    \node[hyper, left=of Sb_f, yshift=1.2em] (Sbm) {$\tilde{\Sigma}_{\mathrm{b},\mu}$};

    \factor[left=of g] {g_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of g_f, yshift=-1.2em] (gs) {$\gamma_\sigma$};
    \node[hyper, left=of g_f, yshift=1.2em] (gm) {$\gamma_\mu$};

    \factor[left=of n] {n_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of n_f, yshift=-1.2em] (ns) {$n_\sigma$};
    \node[hyper, left=of n_f, yshift=1.2em] (nm) {$n_\mu$};

    \factor[left=of a] {a_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of a_f, yshift=-1.2em] (as) {$a_\sigma$};
    \node[hyper, left=of a_f, yshift=1.2em] (am) {$a_\mu$};

@@

    % plates
    \plate {rS} {(R)(S)(sigma)(Shat)} {$\forall R,\sigma_\Sigma^2$};
    \plate {pars} {(R)(S)(sigma)(rb)(Re)(Sb)(g)(n)(a)} {$\forall$ child};

>>>>

    % define nodes
    % define layer 1
    \node[obs] (Shat) {$\log_{10}\hat{\Sigma}$};
    \factor[below=of Shat] {S_f} {right:Norm} {} {};
    \node[det, below=of S_f] (S) {$\log_{10}\Sigma$};
    \node[obs, below=of S] (R) {$\hat{R}$};
    \node[obs, right=of S] (sigma) {$\hat{\sigma_\Sigma}^2$};

    % define layer 2
    \node[latent, left=of S] (Sb) {$\log_{10}{\Sigma}_\mathrm{b}$};
    \node[latent, below=of Sb] (Re) {$R_\mathrm{e}$};
    \node[latent, below=of Re] (rb) {$r_\mathrm{b}$};
    \node[latent, above=of Sb] (g) {$\gamma$};
    \node[latent, above=of g] (n) {$n$};
    \node[latent, above=of n] (a) {$\alpha$};

    % define layer 3
    \factor[left=of rb, xshift=-0.5em] {rb_f} {below:$\mathcal{N}_\mathrm{T}$} {} {};
    \node[hyper, left=of rb_f, yshift=-1.2em] (rbs) {$r_{\mathrm{b},\sigma}$};
    \node[hyper, left=of rb_f, yshift=1.2em] (rbm) {$r_{\mathrm{b},\mu}$};

    \factor[left=of Re, xshift=-0.5em] {Re_f} {below:$\mathcal{N}_\mathrm{T}$} {} {};
    \node[hyper, left=of Re_f, yshift=-1.2em] (Res) {$R_{\mathrm{e},\sigma}$};
    \node[hyper, left=of Re_f, yshift=1.2em] (Rem) {$R_{\mathrm{e},\mu}$};

    \factor[left=of Sb, xshift=-0.5em] {Sb_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of Sb_f, yshift=-2.5em, xshift=-1em] (Sbs) {$\log_{10} \Sigma_{\mathrm{b},\sigma}$};
    \node[hyper, left=of Sb_f, yshift=2.5em, xshift=-1em] (Sbm) {$\log_{10}  \Sigma_{\mathrm{b},\mu}$};

    \factor[left=of g, xshift=-0.5em] {g_f} {below:$\mathcal{N}_\mathrm{T}$} {} {};
    \node[hyper, left=of g_f, yshift=-1.2em] (gs) {$\gamma_\sigma$};
    \node[hyper, left=of g_f, yshift=1.2em] (gm) {$\gamma_\mu$};

    \factor[left=of n, xshift=-0.5em] {n_f} {below:$\mathcal{N}_\mathrm{T}$} {} {};
    \node[hyper, left=of n_f, yshift=-1.2em] (ns) {$n_\sigma$};
    \node[hyper, left=of n_f, yshift=1.2em] (nm) {$n_\mu$};

    \factor[left=of a, xshift=-0.5em] {a_f} {below:$\mathcal{N}_\mathrm{T}$} {} {};
    \node[hyper, left=of a_f, yshift=-1.2em] (as) {$\alpha_\sigma$};
    \node[hyper, left=of a_f, yshift=1.2em] (am) {$\alpha_\mu$};

@@

    % plates
    \plate {rS} {(R)(S)(sigma)(Shat)} {$\forall \hat{R},\hat{\sigma}_\Sigma^2$};
    \plate {pars} {(R)(S)(sigma)(rb)(Re)(Sb)(g)(n)(a)} {$\forall$ child};

### 2023-05-09 papers/paper-1/sections/results.tex

\subsection{Comparison to Previous Work}
\begin{enumerate}
    \item \citet{nasim20} show convergence of eccentricity to follow $\sigma_\mathrm{ecc} \propto 1/\sqrt{N_\star}$
    \item Perform same analysis
    \item Results disagree $\rightarrow$ because of highly radial orbit?? 
\end{enumerate}

>>>>

\subsection{Comparison to Previous Work}
\begin{enumerate}
    \item \citet{nasim20} show convergence of eccentricity to follow $\sigma_\mathrm{ecc} \propto 1/\sqrt{N_\star}$
    \item Perform same analysis
    \item Results disagree $\rightarrow$ because of highly radial orbit?? 
\end{enumerate}
We begin our analysis by confirming the previous result of \citet{nasim20}, that \textit{intra-}simulation eccentricity variance decreases as a function of mass resolution between the SMBH and stellar particles.
For each simulation at a given mass resolution, we take the variance of the eccentricity $\sigma^2_e$ across five orbital periods centred on when the SMBH binary becomes hard, $a=a_\mathrm{h}$. 
We then determine the mean of the variances, and take the square-root of this value, to determine our expected intra-simulation eccentricity spread $\bar{\sigma}_e$.
We fit the same functional form as \citet{nasim20} between $\bar{\sigma}_e$ and $N_\star$, namely:
\begin{equation}
    \bar{\sigma}_e(N_\star;k) = \frac{k}{\sqrt{N_\star + k^2}},
\end{equation}
and find the fit is optimised with a choice of $k=5.31$. 
This is shown in \autoref{fig:nasim}.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{nasim_plot.pdf}
    \caption{\drafting{Intra-simulation binary ecccentricity standard deviation as a function of mass resolution, where the variation is seen to decrease as $\propto 1/\sqrt{N_\star}$. Error bars indicate the $1-\sigma$ spread in $\bar{\sigma}_e$ (note the error bars pertain to the $y$-axis, and not the $x$-axis.)}}
    \label{fig:nasim}
\end{figure}

### 2023-05-09 papers/paper-1/sections/stats.tex

In this work, we have 36 merger configuration families, with each family containing ten perturbed realisations of the same initial state. 
As we only slightly perturb the SMBH phase space coordinates, each \child simulation can be thought of as a sampled observation of the general galaxy-merger configuration for its corresponding family. 
Thus, we are able to employ hierarchical modelling to recover latent parameter distributions for various observable quantities of the merger remnant.

>>>>

In this work, we have eight mass resolution families, with each family containing ten realisations of the same initial state. 
Thus, we are able to employ hierarchical modelling to recover latent parameter distributions for various observable quantities of the merger remnant.

### 2023-05-09 papers/paper-2/sections/numerical.tex
Edits to paper 1

The stellar component, whilst having a spherically-symmetric density distribution, employs the anisotropic Osipkov-Merritt distribution function (DF), from which the phase space is Monte-Carlo sampled using a minimum of $\sim10^6$ particles with a mass of $10^5\,\text{M}_\odot$. 
The Osipkov-Merritt DF has a free parameter $r_a$, termed the anisotropy radius, which determines the degree of radial anisotropy. 
The DF is near-ergodic in the inner regions $r<r_a$, and becomes increasingly radially-biased at larger $r$.
It has the form:
\begin{equation}\label{eq:om}
    f(Q) = \frac{1}{\sqrt{8}\pi^2} \left[ \int_0^Q \dv{\Psi}{\sqrt{Q-\Psi}} \dv{^2\nu_q}{\Psi^2} + \frac{1}{\sqrt{Q}} \left(\dv{\nu_Q}{\Psi}\right)_{\Psi=0}\right],
\end{equation}
where
\begin{equation*}
    \nu_Q(r) \equiv \left(1+\frac{r^2}{r_a^2}\right) \nu(r)
\end{equation*}
and
\begin{equation*}
    Q \equiv \mathcal{E} - \frac{L^2}{2r_a^2}.
\end{equation*}
The parameter $r_a$ is chosen so as to reproduce observed kinematic relations, most notably the $M_\bullet\text{--}\sigma_\star$ relation (see \drafting{figure}), whilst maintaining a positive DF.

@@

>>>>

The stellar component, whilst having a spherically-symmetric density distribution, employs the anisotropic Osipkov-Merritt distribution function (DF), from which the phase space is Monte-Carlo sampled using a minimum of $\sim10^6$ particles with a mass of $10^5\,\text{M}_\odot$. 
The Osipkov-Merritt DF has a free parameter $r_a$, termed the anisotropy radius, which determines the degree of radial anisotropy. 
The DF is near-ergodic in the inner regions $r<r_a$, and becomes increasingly radially-biased at larger $r$.
It has the form:
\begin{equation}\label{eq:om}
    f(Q) = \frac{1}{\sqrt{8}\pi^2} \left[ \int_0^Q \dv{\Psi}{\sqrt{Q-\Psi}} \dv{^2\nu_Q}{\Psi^2} + \frac{1}{\sqrt{Q}} \left(\dv{\nu_Q}{\Psi}\right)_{\Psi=0}\right],
\end{equation}
where
\begin{equation*}
    \nu_Q(r) \equiv \left(1+\frac{r^2}{r_a^2}\right) \nu(r)
\end{equation*}
and
\begin{equation*}
    Q \equiv \mathcal{E} - \frac{L^2}{2r_a^2}.
\end{equation*}
The parameter $r_a$ is chosen so as to reproduce observed kinematic relations, most notably the $M_\bullet\text{--}\sigma_\star$ relation (see \drafting{figure}), whilst maintaining a positive DF.

@@

In typical N-body simulations, one tests the sensitivity of the model by Monte Carlo sampling the phase space distribution function given by \autoref{eq:eddington} a number of times with different random seeds.
This is done to quantify the uncertainty in the model due to the discrete sampling of the phase space of field particles, which induce a Brownian motion to the orbit of the more massive SMBH particle. 
We observe that resampling \autoref{eq:eddington} is in effect displacing the centre of mass (CoM) of the system with the respect to the SMBH.
To avoid the need to run a large number of simulations from the large $3.0\Rvir$ initial separation, we instead \textit{displace the SMBH} with respect to the CoM of the system at a time $t_\mathrm{perturb}$ after the simulation has begun. 
This is done by measuring the Brownian-motion induced displacement in position and velocity space of the SMBH from the CoM for a galaxy in isolation.
After identifying the displacement along each axis to be Gaussian-distributed, we use the variance of this distribution to reposition the SMBH relative to the CoM. 
We thus `recentre' the SMBH on the CoM with a perturbation of $\mathcal{N}_\text{pos}(\mu=0\,\text{pc}, \sigma^2=100\,\text{pc}^2)$ for each position axis, and $\mathcal{N}_\text{vel}(\mu=0\,\text{km\,s}^{-1}, \sigma^2=100\,\text{km}^2\text{\,s}^{-2})$ for each velocity axis. 
In this way, we mimic the effect of Brownian motion for a single run beginning at large separation (the \parent{} run) to generate an arbitrary number of perturbed \child{} runs. 
The perturbation is applied no later than the second-to-last pericentre passage before the SMBH particles form a bound binary system, and ensure the separation of the SMBHs at the time of the perturbation is of the order 1000 larger than the perturbation applied.

### 2023-05-09 papers/paper-eccentricity/sections/numerical.tex

>>>>

\subsection{Galaxy ICs}
We construct our galaxy models to match those of \citet{gualandris22}.
Each galaxy is represented as a stellar-only Dehnen sphere \citep{dehnen93} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed with zero initial velocity.
We test \drafting{X} different mass resolutions by sampling the stellar phase space distribution with varying number of stellar particles: \drafting{list resolutions here}.

\subsection{Merger ICs}
We construct isolated merger initial conditions by placing two galaxies on an elliptical orbit with varying values of eccentricity, at a fixed initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
We follow the same orbital configurations as \citet{gualandris22}, with one additional orbit. 
The chosen eccentricities $e$ are 0.5, 0.7, 0.9, and 0.99.
We ensure that the radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris22}.
An example of each orbital configuration is shown in \autoref{fig:orbits}.
For each orbital configuration, we run \drafting{X} realisations, to account for Brownian motion effects of the SMBH binary caused by the discretised phase space.

\begin{figure*}
    \includegraphics{orbits}
    \caption{\drafting{Orbits of the different merger systems, colour coded according to the initial orbital eccentricity. Note that the duration shown for each orbit is not currently consistent: probably should cut the plot to when the binary semimajor axis reaches some set value}}
    \label{fig:orbits}
\end{figure*}
